---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setlocale, echo=FALSE, results="hide" }
Sys.setlocale("LC_ALL","English")
Sys.setenv(TZ='Europe/Moscow')
```

## Loading and preprocessing the data

```{r loading}

activity_data<-read.csv('activity.csv', colClasses=c("numeric", "character", "integer"))
summary(activity_data)

```

## What is mean total number of steps taken per day?

Now we calculate total number of steps  

```{r totalsteps}

day_steps<-tapply(activity_data$steps, activity_data$date, sum, na.rm=TRUE)

```

Here is the histogram:  
```{r steps_hist}

hist(day_steps, main="Histogram of total steps per day", xlab="steps per day")

```

```{r steps_mean_median}

mean_day_steps<-mean(day_steps)
median_day_steps<-median(day_steps)

```

Mean steps per day is `r mean_day_steps`, median is `r median_day_steps`:

## What is the average daily activity pattern?

Average number of steps per interval between days:

```{r ave_steps_per_interval}

ave_steps<-tapply(activity_data$steps, activity_data$interval, mean, na.rm=TRUE)

```

Activity pattern:

```{r activity}

plot(names(ave_steps), ave_steps, type='l', main='Average number of steps in 5-min interval', xlab='interval', ylab='number of steps')

```

```{r max_interval}
max_step_interval<-as.integer(names(which.max(ave_steps)))
```
Interval `r max_step_interval` contains maximum average number of steps.


## Inputing missing values

```{r miss_val}
total_na<-sum(is.na(activity_data$steps))
```

Total number of missing values is `r total_na`.  
Now we use mean for that interval to fill missing values.

```{r fill_miss_val,cache=FALSE}
# create vector of average values for each interval
ave_int<-sapply(activity_data$interval, function(x) { ave_steps[as.character(x)] })
# create new data set
nomiss_data<-activity_data
# fill missing values
nomiss_data$steps[is.na(activity_data$steps)] <- ave_int[is.na(activity_data$steps)]
# steps per day
nomiss_day_steps<-tapply(nomiss_data$steps, nomiss_data$date, sum)
nomiss_mean<-mean(nomiss_day_steps)
nomiss_median<-median(nomiss_day_steps)
```

Histogram:

```{r nomiss_histogram}
hist(nomiss_day_steps, main="Histogram of total steps per day, missing values filled", xlab="steps per day")
```

Mean steps per day: don't fill missing values: `r mean_day_steps`, fill missing values: `r nomiss_mean`    

Median steps per day: don't fill missing values: `r median_day_steps`, fill missing values: `r nomiss_median`  


## Are there differences in activity patterns between weekdays and weekends?

Now we create new factor in our dataset to indicate weekday or weekend:

```{r weekend, cache=FALSE}
# function to discriminate
is_we<-function(x) { if (weekdays(as.Date(x))=="Saturday" || weekdays(as.Date(x))=="Sunday") as.factor("weekend") else as.factor("weekday") }
# apply to dataset and create new factor variable
suppressWarnings(nomiss_data$is_weekend<-sapply(nomiss_data$date, is_we))
# average by weekend
ave_steps_weekend<-tapply(nomiss_data$steps[nomiss_data$is_weekend=="weekend"], nomiss_data$interval[nomiss_data$is_weekend=="weekend"], mean, na.rm=TRUE)
#average by weekday
ave_steps_weekday<-tapply(nomiss_data$steps[nomiss_data$is_weekend=="weekday"], nomiss_data$interval[nomiss_data$is_weekend=="weekday"], mean)


```

And plot weekdays and weekend average:

```{r weekend_plot, cache=FALSE}
par(mfcol=c(2,1))
plot(as.integer(names(ave_steps_weekend)), ave_steps_weekend, asp=1, type='l', main='weekend', xlab='Interval', ylab='Number of steps')

plot(as.integer(names(ave_steps_weekday)), asp=1, ave_steps_weekday, type='l', main='weekdays', xlab='Interval', ylab='Number of steps')
```

Draw the same using lattice package:

```{r weekend_lattice, cache=FALSE}
library(lattice)

data_ave_steps<-data.frame(interval=c(as.integer(names(ave_steps_weekend)), as.integer(names(ave_steps_weekday))) , steps=c(ave_steps_weekend, ave_steps_weekday), wd=c(rep("weekend",length(ave_steps_weekend) ),rep( "weekdays",length(ave_steps_weekday))))

xyplot(steps ~ interval | wd, data=data_ave_steps, layout=c(1,2), panel=function(x, y, ...){ panel.lines(x,y,...)})                     
                     

```

We can see that there is a difference between weekday and weekend activity patterns - we observe greater mid-day activity on weekend.